Interactive Visualization of Hierarchically Structured Data

### 2017-02-15 treelapse.Rmd
Add a section about the iris data

```{r set_chunk_opts, echo = FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "250px",
  echo = FALSE,
  cache = TRUE
)
```

@@

```{r antibioticoverview, fig.cap = "The two antibiotic time courses are readily apparent even when only highlighting the most abundant taxa."}
knitr::include_graphics("figure/antibiotic_overview.pdf")
```

```{r antibioticbacteroidetes, fig.cap = "Introducing a second box into the timebox display identifies the Bacteroidaceae as a taxon only midly impacted by antibiotics."}
knitr::include_graphics("figure/antibiotic_bacteroidetes.pdf")
```

@@

```{r antibioticfirmicutes, fig.cap = "Zooming into the second antibiotic timecourse and highlighting the Lachnospiraceae and Ruminococcus, we see that the Ruminoccocus took more time to recover to pre-treatment levels."}
knitr::include_graphics("figure/antibiotic_firmicutes.png")
```

@@

```{r ruminococcus, fig.cap = "By hovering over the Ruminoccocus branches, we see that there is a prolonged effect of the antibiotics time courses more or less uniformly across the lower taxonomic orders, with the except of "}
knitr::include_graphics("figure/ruminococcus.pdf")
```

@@

```{r verrucomicrobiae, fig.cap = "The subtree averages aggregation brings attention to the Verrucomicrobiae, which though only present as a few species, are each rather abundant. In particular, they seem to increase after the first antibiotic time course."}
knitr::include_graphics("figure/verrucomicrobiae.pdf")
```

@@

```{r pretermcsts, fig.cap = "The increased diversity among samples in CST 4 is represented by the relatively larger contribution of red edges to branches outside of the Firmicutes."}
knitr::include_graphics("figure/preterm_csts.pdf")
```

```{r pretermcstslacto, fig.cap = "Zooming into the Lactobacillaceae family, we notice that the difference between the remaining four CSTs is related to which types of Lactobacillus are most prominent."}
knitr::include_graphics("figure/preterm_csts_lacto.pdf")
```

```{r pretermpreterm, fig.cap = "Samples with high levels of phyla other than Firmicutes appear to be related to preterm births."}
knitr::include_graphics("figure/preterm_preterm.pdf")
```

@@

We next consider an application unrelated to the microbiome, but with
relatively clear hierarchical structure. Our data are downloaded from
Zillow^[[http://www.zillow.com/research/data/](http://www.zillow.com/research/data/)],
and give the Zillow Home Value Indexes at the neighborhood level,
across the country, computed monthly between 1996 and 2016. In our
display, we have taken the base 10 log of these indexes. As our
hierarchical structure, we use each neighborhood's assignment to
state, regional, county, and city levels. We represent each of these
coarser spatial categories using the average of all neighborhoods
contained in them. We have filtered down to the 890 neighborhoods in
California; rendering more neighborhoods while keeping all 246
timepoints causes the interface to lag^[See the Section
\@ref(conclusion) for potential optimizations, however.].

@@

```{r zillowcities, fig.cap = "California home prices at the city level, between 1996 and 2016. The effect of the 2008 recession is clear, and we have hovered over the San Bernardino series, to display the identity of one of the cities most strongly affected by the recession."}
knitr::include_graphics("figure/zillow_cities.pdf")
```

```{r zillowmiddlepre, fig.cap = "Neighborhoods with mid-range home prices before the recession are selected here. Note that the collection of series seems to widen after 2008 -- we are interested in whether there are reliable predictors of these alternate trajectories, given their similar starting points."}
knitr::include_graphics("figure/zillow_middle_pre.pdf")
```

```{r zillowmiddleup, fig.cap = "Among those neighborhoods with mid-range prices before the recession, we have selected those that recovered more rapidly. These appear to be located mainly in Los Angeles and San Diego counties."}
knitr::include_graphics("figure/zillow_middle_up.pdf")
```

```{r zillowmiddledown, fig.cap = "Rather than selecting mid-range series that recovered quickly, we can isolate those whose prices remained depressed after the recession. These seem mostly to be located in Central California and the East San Francisco Bay Area."}
knitr::include_graphics("figure/zillow_middle_down.pdf")
```

```{r zillowsf, fig.cap = "To study teh range in home prices within San Francisco County, we can select the associated nodes using a treebox."}
knitr::include_graphics("figure/zillow_sf.pdf")
```

@@

```{r working, fig.cap = "The two peaks at rush hour distinguish weekday series from the rest."}
knitr::include_graphics("figure/working.pdf")
```

```{r weekend, fig.cap = "Unlike weekday demand, weekend demand is unimodal. The few weekday series with unimodal series seem to be associated with holidays. This is the case for New Years' Eve, which is currently hovered over in the tree."}
knitr::include_graphics("figure/weekend.pdf")
```

```{r weekday2011, fig.cap = "Weeday demand appears larger in 2012 than 2011 -- compare with the next figure."}
knitr::include_graphics("figure/weekday_2011.pdf")
```

```{r weekday2012, fig.cap = "Weekday demand increased in 2012 -- compare with the previous figure."}
knitr::include_graphics("figure/weekday_2012.pdf")
```

```{r warmweekend, fig.cap = "The samples with highest night demand tend to fall on warm weekends."}
knitr::include_graphics("figure/warm_weekend.pdf")
```

```{r bikesankey, fig.cap = "We can interpret the regression tree using an interactive DOI representation."}
knitr::include_graphics("bike_sankey.png")
```

@@

Potential sections to write

* describe extension from svg to canvas
* give example of parallel coordinates / hierarchical clustering?
* explain incorporation into / give example of shiny app?

>>>>

```{r set_chunk_opts, echo = FALSE}
library("knitr")
opts_chunk$set(
  fig.align = "center",
  out.width = "250px",
  echo = FALSE,
  cache = TRUE
)
```

@@

```{r antibioticoverview, fig.cap = "The two antibiotic time courses are readily apparent even when only highlighting the most abundant taxa."}
include_graphics("figure/antibiotic_overview.pdf")
```

```{r antibioticbacteroidetes, fig.cap = "Introducing a second box into the timebox display identifies the Bacteroidaceae as a taxon only midly impacted by antibiotics."}
include_graphics("figure/antibiotic_bacteroidetes.pdf")
```

@@

```{r antibioticfirmicutes, fig.cap = "Zooming into the second antibiotic timecourse and highlighting the Lachnospiraceae and Ruminococcus, we see that the Ruminoccocus took more time to recover to pre-treatment levels."}
include_graphics("figure/antibiotic_firmicutes.png")
```

@@

```{r ruminococcus, fig.cap = "By hovering over the Ruminoccocus branches, we see that there is a prolonged effect of the antibiotics time courses more or less uniformly across the lower taxonomic orders."}
include_graphics("figure/ruminococcus.pdf")
```

@@

```{r verrucomicrobiae, fig.cap = "The subtree averages aggregation brings attention to the Verrucomicrobiae, which though only present as a few species, are each rather abundant. In particular, they seem to increase after the first antibiotic time course."}
include_graphics("figure/verrucomicrobiae.pdf")
```

@@

```{r pretermcsts, fig.cap = "The increased diversity among samples in CST 4 is represented by the relatively larger contribution of red edges to branches outside of the Firmicutes."}
include_graphics("figure/preterm_csts.pdf")
```

```{r pretermcstslacto, fig.cap = "Zooming into the Lactobacillaceae family, we notice that the difference between the remaining four CSTs is related to which types of Lactobacillus are most prominent."}
include_graphics("figure/preterm_csts_lacto.pdf")
```

```{r pretermpreterm, fig.cap = "Samples with high levels of phyla other than Firmicutes appear to be related to preterm births."}
include_graphics("figure/preterm_preterm.pdf")
```

@@

We next consider an application unrelated to the microbiome, but with
relatively clear hierarchical structure. Our data are downloaded from
Zillow^[[http://www.zillow.com/research/data/](http://www.zillow.com/research/data/)],
and give the Zillow Home Value Indexes at the neighborhood level,
across the country, computed monthly between 1996 and 2016. In our
display, we have taken the base 10 log of these indexes. As our
hierarchical structure, we use each neighborhood's assignment to
state, regional, county, and city levels. We represent each of these
coarser spatial categories using the average of all neighborhoods
contained in them. We have filtered down to the 890 neighborhoods in
California; rendering more neighborhoods while keeping all 246
timepoints causes the interface to lag^[See Section
\@ref(conclusion) for potential optimizations, however.].

@@

```{r zillowcities, fig.cap = "California home prices at the city level, between 1996 and 2016. The effect of the 2008 recession is clear, and we have hovered over the San Bernardino series, to display the identity of one of the cities most strongly affected by the recession."}
include_graphics("figure/zillow_cities.pdf")
```

```{r zillowmiddlepre, fig.cap = "Neighborhoods with mid-range home prices before the recession are selected here. Note that the collection of series seems to widen after 2008 -- we are interested in whether there are reliable predictors of these alternate trajectories, given their similar starting points."}
include_graphics("figure/zillow_middle_pre.pdf")
```

```{r zillowmiddleup, fig.cap = "Among those neighborhoods with mid-range prices before the recession, we have selected those that recovered more rapidly. These appear to be located mainly in Los Angeles and San Diego counties."}
include_graphics("figure/zillow_middle_up.pdf")
```

```{r zillowmiddledown, fig.cap = "Rather than selecting mid-range series that recovered quickly, we can isolate those whose prices remained depressed after the recession. These seem mostly to be located in Central California and the East San Francisco Bay Area."}
include_graphics("figure/zillow_middle_down.pdf")
```

```{r zillowsf, fig.cap = "To study teh range in home prices within San Francisco County, we can select the associated nodes using a treebox."}
include_graphics("figure/zillow_sf.pdf")
```

@@

```{r working, fig.cap = "The two peaks at rush hour distinguish weekday series from the rest."}
include_graphics("figure/working.pdf")
```

```{r weekend, fig.cap = "Unlike weekday demand, weekend demand is unimodal. The few weekday series with unimodal series seem to be associated with holidays. This is the case for New Years' Eve, which is currently hovered over in the tree."}
include_graphics("figure/weekend.pdf")
```

```{r weekday2011, fig.cap = "Weeday demand appears larger in 2012 than 2011 -- compare with the next figure."}
include_graphics("figure/weekday_2011.pdf")
```

```{r weekday2012, fig.cap = "Weekday demand increased in 2012 -- compare with the previous figure."}
include_graphics("figure/weekday_2012.pdf")
```

```{r warmweekend, fig.cap = "The samples with highest night demand tend to fall on warm weekends."}
include_graphics("figure/warm_weekend.pdf")
```

```{r bikesankey, fig.cap = "We can interpret the regression tree using an interactive DOI representation."}
include_graphics("figure/bike_sankey.pdf")
```

### Hierarchically Clustering the Iris Data {#iris}

Each of the timebox tree and treebox examples presented so far have focused on
data with a clear time component. We note however that these techniques could
alternatively be applied to high-dimensional data, via the use of parallel
coordinates [@inselberg1991parallel]. The usual parallel coordinates challenges
remain, namely selecteing scales for and an ordering across the different
coordinates, but the linking and focus-plus-context ideas can still be employed
in this setting. Here we provide a simple implementation of this idea, using the
famous Iris data, inspired by the hierarchical clustering vignette in
[@galili2015dendextend].

The Iris data set is a collection of 150 measurements of four variables --
widths and lengths of Iris sepals and petals [@fisher1936use]. There are 50
samples for each of three species of Iris, \textit{setosa}, \textit{versicolor}
and \textit{virgnica}. As in [@galili2015dendextend], we hierarchical cluster
samples\footnote{We use complete linkage and the euclidean distance.}. Our
contribution is to display the resulting hierarchy together with a parallel
coordinates view of the raw data, see Figure \@ref(fig:iristimebox).

```{r iristimebox, fig.cap = "A toy application to the Iris data gives an example of how linking in treelapse can be applied to combine hierarchical clustering and parallel coordinates views."}
include_graphics("iris_timebox.pdf")
```

From Figure \@ref(fig:iristimebox), the fact that the highest split in the
hierarchical clustering distinguishes those samples with small petal lengths
from the rest. Upon inspecting the associated IDs, we find that this cluster
includes all the samples from setosa. The similarity between versicolor and
virginica is reflected in their proximity in the clustering. Further, applying a
treebox view to scan the subtrees associated with these two species reveals the
more subtle differences in sepal and petal lengths that can be used to
distinguish them.

@@
